---
title: "Project STAT4540"
output:
  word_document: default
  html_document:
    df_print: paged
date: "2023-11-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(class)
library(caret)
library(FNN)
library(MASS)
library(e1071)
library(stats)  # for hclust
library(cluster)  # for kmeans
library(glmnet)
library(pls)
library(leaps)
```

```{r}
# Load data
train_df <- read.csv("C:/Users/johnC/Downloads/train_mnist.csv")
test_df <- read.csv("C:/Users/johnC/Downloads/test_mnist.csv")
```
(Q1)

[1]
```{r 11}

# Define the predictor and response variables
x_train <- as.matrix(train_df[, 1:784])
y_train <- as.factor(train_df[, "y"])
x_test <- as.matrix(test_df[, 1:784])
y_test <- as.factor(test_df[, "y"])

# Create a data frame to store results
result_df <- data.frame(Flexibility = numeric(), ErrorRate = numeric())

# Define a range of K values
k_values <- c(1, 2, 3, 10, 100, 500, 750, 1000)

# Iterate over different K values
for (k in k_values) {
    # Train K-NN model
    knn_model <- knn(train = x_train, test = x_test, cl = y_train, k = k)
    
    # Make predictions on test data
    predictions <- as.factor(knn_model)
    
    # Calculate the confusion matrix
    confusion_matrix <- confusionMatrix(predictions, y_test)
    
    # Extract error rate and store in result_df
    error_rate <- 1 - confusion_matrix$overall["Accuracy"]
    result_df <- rbind(result_df, data.frame(Flexibility = 1/k, ErrorRate = error_rate))
}

# Plot the test error rate curve
plot(result_df$Flexibility, result_df$ErrorRate, type = "b", 
     xlab = "Flexibility (1/K)", ylab = "Test Error Rate",
     main = "K-NN: Test Error Rate Curve")

# Identify the optimal Flexibility with the minimum error rate
optimal_flexibility <- result_df$Flexibility[which.min(result_df$ErrorRate)]
cat("Optimal Flexibility:", optimal_flexibility, "\n")
cat("Best Error Rate:", min(result_df$ErrorRate), "\n")
```
[2]
```{r 12}
# True Positive Rate Calculation
knn_tpr <- sapply(k_values, function(k) {
  knn_pred <- knn(train = train_df[, 1:784], test = test_df[, 1:784], cl = train_df$y, k = k)
  sapply(0:9, function(dig) mean(knn_pred[test_df$y == dig] == dig))
})

# Plotting Class-Specific TPR Against Flexibility
matplot(t(matrix(knn_tpr, nrow = 10)), type = "b", pch = 1, col = 1:10,
        xlab = "Flexibility (1/K)", ylab = "True Positive Rate",
        main = "Class-Specific TPR vs Flexibility")

legend("topright", legend = as.character(0:9), col = 1:10, pch = 1, title = "Digit", cex = 0.8)

```
[3]
```{r 13}
# 3. LDA 

lda_fit <- lda(y ~ ., data = train_df)
lda_pred <- predict(lda_fit, test_df[,1:784])$class
lda_err <- mean(lda_pred != test_df$y)
lda_tpr <- t(sapply(0:9, function(dig) mean(lda_pred[test_df$y==dig] == dig)))

lda_err
lda_tpr

# Discriminant function is linear combination of predictors 
# Parameters are within class means (10), common covariance matrix (784x784)
```
[4]
```{r 14}
# 4. QDA
qda_fit <- qda(y ~ ., data = train_df)
qda_pred <- predict(qda_fit, test_df[,1:784])$class
qda_err <- mean(qda_pred != test_df$y)
qda_tpr <- t(sapply(0:9, function(dig) mean(qda_pred[test_df$y==dig] == dig)))

qda_err
qda_tpr


# Parameters are class-specific means (10x784) and covariances (10x784x784) 
```
[5]

The \texttt{e1071} package estimates the class-specific posterior probabilities using a Bayesian approach with asymptotic likelihood estimates. To put it more precisely:

By utilizing maximum likelihood estimation (MLE) to ascertain the ratio of training examples in each class, the method computes the class prior probabilities.

Based on each class's unique conditions, the algorithm determines its density. $p(x|y=c_k)$ is the conditional probability of $x$ given $y$ equals $c_k$. It uses kernel density estimation to obtain a non-parametric estimate of the density.

It then computes the posterior probabilities using Bayes' rule. $p(y=c_k|x)$ is the conditional probability that $y$ equals $c_k$ given $x$.

```{r 15}
# 5. Naive Bayes

nb_fit <- naiveBayes(y ~ ., data = train_df)
nb_pred <- predict(nb_fit, test_df[,1:784])
nb_err <- mean(nb_pred != test_df$y)
nb_tpr <- t(sapply(0:9, function(dig) mean(nb_pred[test_df$y==dig] == dig)))

nb_err
nb_tpr

# Estimates class probabilities by relative frequencies
```
[6]
```{r 16}
# 6. Comparison
methods <- c("KNN", "LDA", "QDA", "Naive Bayes")
err_rates <- c(min(result_df$ErrorRate), lda_err, qda_err, nb_err)

barplot(err_rates, names.arg=methods, ylab="Error Rate")

# LDA has lowest error rate, followed by QDA
```

Q2.

1. 

\begin{itemize}

\item Extract images of the digits 1 and 8 from both the training and test sets

\item Convert the provided information to matrix format

\item Perform data centering by subtracting the mean of every feature

\item Scale the data by dividing by the standard deviation of each feature

\textbf Principal Component Analysis (PCA) Optimization:

\item Determine which principal components (PCs) capture the most variance in the dataset

\item The first two principal components, $\psi_1$ and $\psi_2$, represent the directions of highest variability in the feature space

\item The principal component (PC) scores represent the position values of the data points on the new basis

\textbf Dimensions:

\item The sample size of the training data is the number of images used to represent digits 1 and 8

\item The dimensions of the principal component (PC) scores are determined by the number of samples and number of PCs

\item The dimensions of the PC directions are determined by the number of PCs and number of features

\end{itemize}


```{r}
idx8 <- which(train_df$y == 8)
idx1 <- which(train_df$y == 1)
xtrain8 <- as.matrix(train_df[idx8, 1:784])
xtrain1 <- as.matrix(train_df[idx1, 1:784])
id8 <- which(test_df$y == 8)
id1 <- which(test_df$y == 1)
xtest8 <- as.matrix(test_df[id8, 1:784])
xtest1 <- as.matrix(test_df[id1, 1:784])
```

2. 
```{r 22}
pcout8 <- prcomp(xtrain8)
pcout1 <- prcomp(xtrain1)
```

The elements of pcout8 and pcout1 are as follows, in accordance with their respective definitions:  

Standard deviation (sdev) for every principal component

rotation: The principal component loading vectors

Scale and center: The data centering and scaling methods used

x: Rotated values of the data following PCA transformation


3.

The prcomp() function gives two outputs, pcout8$center and pcout8$scale. These are used to preprocess the training data, centering and scaling it before Principal Component Analysis (PCA). 

In order to center the training data, the average values of each column were subtracted, and this is represented by the variable "pcout8$center". 

The variable pcout8$scale indicates whether any scaling was implemented in addition to centering.  The training data was scaled by dividing it by the standard deviations of each column, if the condition is True. 

```{r 23}
# Apply the same centering and scaling to test data as in training data
xtest8_centered_scaled <- scale(xtest8, center = pcout8$center, scale = pcout8$scale)
xtest1_centered_scaled <- scale(xtest1, center = pcout8$center, scale = pcout8$scale)
```

4. 
\begin{align*}
\boldsymbol{\varphi}_1 &\text{: First principal component direction (loading vector)} \\
\boldsymbol{\varphi}_2 &\text{: Second principal component direction (loading vector)}
\end{align*}

The variables $\boldsymbol{\varphi}_1$ and $\boldsymbol{\varphi}_2$ represent the first two columns of the rotation matrix $\mathbf{pcout8$rotation}$. 
These vectors denote the directions within the original pixel space that most effectively capture the variance in the given data. The principal axes, determined as the directions along which the data exhibits the highest variability, are indicated by $\boldsymbol{\varphi}_1$ and $\boldsymbol{\varphi}_2$.

Geometrically, $\boldsymbol{\varphi}_1$ and $\boldsymbol{\varphi}_2$ define two orthogonal axes that form a new coordinate system for the pixel space. Projecting the original data onto $\boldsymbol{\varphi}_1$ and $\boldsymbol{\varphi}_2$ results in new features that preserve the directions of maximum variance while filtering out noise. Thus, these loading vectors point along the directions of greatest variability in the image data.

5.
\begin{align*}
\boldsymbol{x}_1^{\top} &\text{: First observation in original pixel space (first row of }\mathbf{xtrain8}) \\
\boldsymbol{z}_1^{\top} &\text{: First observation in principal components space (first row of }\mathbf{pcout8$x})
\end{align*}

The variable $\boldsymbol{z}$ represents the projection of the original variable $\boldsymbol{x}$ onto the principal components coordinate system. Specifically, $\boldsymbol{z}_{1}^{\top}$ provides the coordinates of the first observation with respect to the directions of maximum variance (principal components). By transforming from $\boldsymbol{x}$ to $\boldsymbol{z}$, the data is expressed in a more concise form that captures the important variability characteristics.

Geometrically, $\boldsymbol{z}_{1}^{\top}$ gives the coordinates of the first observation on the new orthogonal axes defined by the principal directions. This allows the high-dimensional pixel data to be effectively summarized in the lower-dimensional space spanned by the principal components. The key information concerning the variation patterns in $\boldsymbol{x}$ is preserved in $\boldsymbol{z}$.



6. 

pcout8$sdev contains the standard deviations of the principal components.


7.

```{r 27}
cum_var_explained_8 <- cumsum(pcout8$sdev^2) / sum(pcout8$sdev^2)
cum_var_explained_1 <- cumsum(pcout1$sdev^2) / sum(pcout1$sdev^2)

# Plotting
plot(cum_var_explained_8, type = "l", col = "blue", xlab = "Number of Principal Components",
     ylab = "Cumulative Proportion of Variance Explained", main = "Cumulative Variance Explained by PCs")
lines(cum_var_explained_1, type = "l", col = "red")
legend("bottomright", legend = c("pcout8", "pcout1"), col = c("blue", "red"), lty = 1)
```
Using pcout1 allows for a greater explanation of variance with a smaller number of principal components, while pcout8 utilizes a larger number of principal components that explain less variance.


8.

```{r 28p}
# Assuming xtest8 and xtest1 are your centered-and-scaled test data
# Assuming pcout8 and pcout1 are your PCA results for xtrain8 and xtrain1


# Number of loadings to use (M = 1 to 500)
M_values <- 1:500
err8 = numeric(500)
err1 = numeric(500)
# Reconstruct images for each M value
for (M in M_values) {
  # Reconstruct for xtest8
  load8 = pcout8$rotation[, 1:M]
  z8 = xtest8_centered_scaled %*% load8 
  reconstructed_xtest8 <- z8 %*% t(load8)
  err8[M] = sum((xtest8_centered_scaled - reconstructed_xtest8)^2) / nrow(xtest8_centered_scaled) 
  
    # Reconstruct for xtest1
  load1 = pcout1$rotation[, 1:M]
  z1 = xtest1_centered_scaled %*% load1 
  reconstructed_xtest1 <- z1 %*% t(load1)
  err1[M] = sum((xtest1_centered_scaled - reconstructed_xtest1)^2) / nrow(xtest1_centered_scaled) 

  # Now, you have reconstructed_xtest8 and reconstructed_xtest1 for each M value
  # You can use or analyze these reconstructed images as needed
}

# Heatmap color palette 
cols <- rev(heat.colors(256))

par(mfrow=c(1, 2))

# Digit 1 plots
image(matrix(xtest1_centered_scaled[1,], ncol = 28), col = cols, 
      main = "Original xtest1")

image(matrix(reconstructed_xtest1[1,], ncol = 28), col = cols,  
      main = "Reconstructed xtest1")
      
# Digit 8 plots 
image(matrix(xtest8_centered_scaled[1,], ncol = 28), col = cols,
      main = "Original xtest8")

image(matrix(reconstructed_xtest8[1,], ncol = 28), col = cols,
      main = "Reconstructed xtest8")
```


9. 

```{r 29}

# Plot reconstruction errors
plot(M_values, err8, type = "l", col = "blue", xlab = "Number of Principal Components (M)", ylab = "Reconstruction Error", main = "Reconstruction Error vs. M for xtest8")
lines(M_values, err1, type = "l", col = "red")
legend("topright", legend = c("xtest8", "xtest1"), col = c("blue", "red"), lty = 1)

```



Q3.

[i]
```{r 31}
# Assuming xtrain8 and xtrain1 are matrices containing raw training images for digits 8 and 1

# (i) Perform hierarchical clustering
X <- rbind(xtrain8, xtrain1)
hc.comp <- hclust(dist(X), method = "complete")
hc.avg <- hclust(dist(X), method = "average")
hc.sgl <- hclust(dist(X), method = "single")

# Plot dendrograms
par(mfrow = c(1, 3))
plot(hc.comp, main = "Complete Linkage")
plot(hc.avg, main = "Average Linkage")
plot(hc.sgl, main = "Single Linkage")
```
The greatest distance between any two points in the distinct clusters is considered in complete linkage. This increases its resistance to outliers, or data points that deviate significantly from the primary cluster. It has a tendency to form compact and balanced clusters by considering the most dissimilar pair of points when deciding how to combine clusters. 

[ii]
```{r}
# (ii) Cut tree for 2 clusters using complete linkage
cutree_comp <- cutree(hc.comp, k = 2)

# Check if the clusters correspond to digits 8 and 1
table(cutree_comp, rep(c(8, 1), each = length(cutree_comp) / 2))
```
Yes. They do correspond to the digits 1 and 8. 

[iii]
```{r}
# (iii) Repeat using PC scores
pcout8 <- prcomp(xtrain8)
pcout1 <- prcomp(xtrain1)

# Skip the check and conversion, as princomp returns a matrix
Z <- rbind(pcout8$x, pcout1$x)[, 1:3]

# Check the structure of Z and convert it to a matrix if needed
str(Z)

# If Z is a data frame, convert it to a matrix
if (is.data.frame(Z)) {
  Z <- as.matrix(Z)
}

# Scale Z with centering and scaling
Z_scaled <- scale(Z, center = TRUE, scale = TRUE)


# Perform hierarchical clustering on scaled Z
hc.comp_Z_scaled <- hclust(dist(Z_scaled), method = "complete")
cutree_comp_Z_scaled <- cutree(hc.comp_Z_scaled, k = 2)

# Check if the clusters correspond to digits 8 and 1
table(cutree_comp_Z_scaled, rep(c(8, 1), each = length(cutree_comp_Z_scaled) / 2))

# Compare results and provide justification
```



(2) 2-Means clustering
 (i) Describe the 2-Means optimization problem
 Dissimilarity measure: Euclidean distance
 Objective: Minimize the sum of squared distances between data points and their assigned cluster centers
 Variables: Cluster centers
 
[ii]
```{r 32}
# (ii) Perform 2-Means on X, Z, and scaled Z with 3 random initializations
set.seed(42)  # for reproducibility
kmeans_X <- kmeans(X, centers = 2, nstart = 3)
kmeans_Z <- kmeans(Z, centers = 2, nstart = 3)
kmeans_Z_scaled <- kmeans(Z_scaled, centers = 2, nstart = 3)
```

[iii]
```{r 33}
# (iii) Compare 2-Means approaches
# Check which approach is better at identifying the cluster with the correct digit label
# Create true labels for comparison
true_labels <- rep(c(8, 1), c(length(xtrain8), length(xtrain1)))

# Compare results and provide justification
table(kmeans_X$cluster)
table(kmeans_Z$cluster)
table(kmeans_Z_scaled$cluster)
```
The initial placement of cluster centroids has a significant impact on the k-means algorithm. Changes to the original cluster centers can yield different solutions.
Performing several random initializations—three in this instance—increases the likelihood of finding a global minimum for the sum of squared distances.





Q4. 

```{r}
# Assuming you have already loaded your data
train_data <- read.csv("C:/Users/johnC/Downloads/reg_train.csv")
test_data <- read.csv("C:/Users/johnC/Downloads/reg_test.csv")
```
[1]. 

Best subset selection is computationally expensive for high-dimensional data

2. Least Squares
```{r 42}
lm_model <- lm(y ~ ., data = train_data)
test_predictions <- predict(lm_model, newdata = test_data)
test_mse_ls <- mean((test_predictions - test_data$y)^2)
```

 3. Forward Selection
```{r 43}
# Set the range of nvmax values
nvmax_values <- 1:199

# Perform forward selection and compute test MSE for each nvmax
mse_values <- sapply(nvmax_values, function(nvmax) {
  forward_model <- regsubsets(y ~ ., data = train_data, method = "forward", nvmax = nvmax)
  
  summary_model <- summary(forward_model)
  best_formula <- names(which.max(summary_model$adjr2))
  
 selected_vars <- c("y", best_formula)

selected_data <- train_data[, selected_vars, drop = FALSE]
  
  best_lm <- lm(y ~ ., data = selected_data)
  
  test_predictions_f <- predict(best_lm, newdata = test_data)
  test_mse_f <- mean((test_predictions_f - test_data$y)^2)
  
  return(test_mse_f)
})

# Find the index of the minimum test MSE
best_nvmax_index <- which.min(mse_values)

# Get the corresponding best nvmax value
best_nvmax <- nvmax_values[best_nvmax_index]

# Print the results
cat("Best nvmax value:", best_nvmax, "\n")
cat("Corresponding test MSE:", mse_values[best_nvmax_index], "\n")

```

 4. Backward Selection
```{r 44}
# Backward selection 

mse_values <- numeric(length = length(nvmax_values))

for (i in seq_along(nvmax_values)) {
  nvmax <- nvmax_values[i]
  
  backward_model <- regsubsets(y ~ ., data = train_data, nvmax = nvmax, method = "backward")
  
  summary_model <- summary(backward_model)
  best_formula <- names(which.max(summary_model$adjr2))
  
  selected_vars <- c("y", best_formula)
  selected_data <- train_data[, selected_vars, drop = FALSE]
  
  best_lm <- lm(y ~ ., data = selected_data)
  
  test_predictions_b <- predict(best_lm, newdata = test_data)
  test_mse_b <- mean((test_predictions_b - test_data$y)^2)
  
  mse_values[i] <- test_mse_b
}

# Find best nvmax
best_nvmax_index <- which.min(mse_values)
best_nvmax <- nvmax_values[best_nvmax_index]

# Print the results
cat("Best nvmax:", best_nvmax, "\n")
cat("Minimum Test MSE:", mse_values[best_nvmax_index], "\n")

```

5. Compare Forward and Backward Selection

For backward selection to work, there must be more samples (n) than variables (p), so that the whole model can be fit. Forward selection, on the other hand, can be used even when n is less than p, and it is the only subset method that works when p is very big. We have n = 200 and p = 500 in this case. Because of this, backward selection is not the best method because all the predictors have to be used in the first round, which is a nxp situation. So, in this case, forward subset selection is the better choice.


6. Ridge and Lasso Regression

Ridge Regression

Minimize \[ \sum_{i=1}^{n} (y_i - \beta_0 - \sum_{j=1}^{p} x_{ij} \beta_j)^2 + \lambda \sum_{j=1}^{p} \beta_j^2 \]

Lasso Regression

Minimize \[ \sum_{i=1}^{n} (y_i - \beta_0 - \sum_{j=1}^{p} x_{ij} \beta_j)^2 + \lambda \sum_{j=1}^{p} |\beta_j| \]


The impact of the parameter \lambda on variance and bias is investigated within the framework of Ridge Regression.

The regularization term gains importance as the value of \lambda rises.

Greater contraction of coefficients towards zero is the outcome of increasing the value of \lambda.

Although the variance is decreasing, the bias is increasing.

Advantageous when multiple predictors exhibit multicollinearity.


Lasso Regression:

The regularization term gains importance as the value of \lambda rises.

A sparser model is produced by raising the value of \lambda because it makes some coefficients exactly zero.

As the degree of variance decreases, bias becomes more prevalent.

Advantageous for feature selection since it typically selects a subset of important predictors.


# Ridge
```{r 46}
ridge_model <- cv.glmnet(as.matrix(train_data[, -ncol(train_data)]), train_data$y, alpha = 0)
best_lambda_ridge <- ridge_model$lambda.min
# Lasso
lasso_model <- cv.glmnet(as.matrix(train_data[, -ncol(train_data)]), train_data$y, alpha = 1)
best_lambda_lasso <- lasso_model$lambda.min
best_lambda_lasso
```

7. Ridge Regression Test MSE
```{r 47}
ridge_test_predictions <- predict(ridge_model, newx = as.matrix(test_data[, -ncol(test_data)]), s = best_lambda_ridge)
test_mse_ridge <- mean((ridge_test_predictions - test_data$y)^2)
non_zero_coef_ridge <- sum(coef(ridge_model, s = best_lambda_ridge) != 0)
non_zero_coef_ridge
test_mse_ridge

```

 8. Lasso Regression Test MSE
```{r 48}
lasso_test_predictions <- predict(lasso_model, newx = as.matrix(test_data[, -ncol(test_data)]), s = best_lambda_lasso)
test_mse_lasso <- mean((lasso_test_predictions - test_data$y)^2)
non_zero_coef_lasso <- sum(coef(lasso_model, s = best_lambda_lasso) != 0)
non_zero_coef_lasso
test_mse_lasso
```

 9. PCR
```{r 49}
# Fit PCR model on the training set
pcr_model <- pcr(y ~ ., scale = TRUE, data = train_data, validation = "CV")

# Get the cross-validated RMSE values for different numbers of components
cv_mse_pcr <- MSEP(pcr_model, newdata = test_data)$val

# Find the optimal number of components (M) based on minimum RMSE
optimal_M_pcr <- which.min(cv_mse_pcr)

# Print results for PLS
print(cv_mse_pcr)
print(optimal_M_pcr)


```

 10. PLS
```{r 410}
pls_cv <- plsr(y ~ ., scale = TRUE, data = train_data, validation = "CV")

# Get the cross-validated RMSE values for different numbers of components
cv_mse_pls <- MSEP(pls_cv, newdata = test_data)$val

# Find the optimal number of components (M) based on minimum RMSE
optimal_M_pls <- which.min(cv_mse_pls)

# Print results for PLS
print(cv_mse_pls)
print(optimal_M_pls)
```
 11. Compare models

Bias-Variance Tradeoff Using Various Techniques:

Principle Component Regression, or PCR:

Lower overall test mean square error (MSE) could result from PCR's capacity to reduce variance by removing multicollinearity and capturing essential variance.

Partial Least Squares, or PLS:

By utilizing data from both predictors and responses, it balances variance reduction and bias, but because it depends so heavily on the response variable, bias may be added.

Ridge Regression:

By introducing a penalty to the OLS objective, ridge regression lowers variance and shrinks coefficients toward zero. Regularization may lead to an increase in bias, but it effectively prevents overfitting, which reduces variance.

Forward Selection:

By gradually adding predictors, forward selection may raise the complexity and variance of the model. Although it can help the model fit the dataset, overfitting could result in higher test MSE and more variance.

Lasso Regression:

By forcing some of the coefficients to zero, Lasso regression reduces the complexity of the model while performing feature selection. Although Lasso is useful for selecting features, it may exacerbate bias by excluding features that could be important, which could affect prediction accuracy.

PCR Selection Considering the Bias-Variance Tradeoff:

PCR with the lowest test mean square error would be the best prediction model for our dataset.

Advantages of PCR:

By converting predictors into uncorrelated components, PCR reduces the possibility of overfitting by addressing multicollinearity. It focuses on capturing important variance while lowering dimensionality, striking a compromise between variance reduction and bias.


